---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# frast

<!-- badges: start -->
<!-- badges: end -->

`frast` is a prototype / proof-of-concept of two things:
 
1)  An R-to-Fortran transpiler
2)  A just-in-time (JIT) compiler for R functions

 
Currently it can take an R function, and translate it to a compiled function (by way of transpiling it to Fortran). Right now it requires users to specify a variable type manifest for the R function, but that shouldn't be necessary once the JIT piece is fleshed out a little more. 

Right now R functions are translated to Fortran subroutines, which are then compiled as a stand-alone Fortran 2018 module. Compilation is performed using gfortran. The compiled shared object is then dynamically loaded using `dyn.load`, and the compiled function in the loaded object can then be evaluated using `RFI::.ModernFortran`. 

The next logical step in this project is to swap out the `gfortran` system calls to `flang` calls that emit LLVM IR bytecode, then use the LLVM C API to evaluate the functions. This would get around the limitations inherent in using `dyn.load`, and should give a nice speed boost too.

Once the llvm piece is in place, then it would might be a good time to work on
removing the need for explicit type manifests in R. (note-to-self: see if it's
easy to recylce the implementation in `package:memoise`). A solid approach would
be to add a check before `eval`ing the R function, to check if a compiled
version of that function matching the calling arguments signature is available.
If not, launch a background task compiling (so it will be available next time
the function is evaluated) but proceed with evaluating the R function as normal.
This will require forcing of all promises, but that should be a small price to
pay for the speed boost. 

## Example

This is a basic example 

```{r}
library(frast)
library(RFI)

addone <- function(x) {
  manifest(Var(x, "d", shape = ":"))
  x = x + 1
}
```



```{r}
(ptr <- load_so(addone))
RFI::.ModernFortran(ptr, array(1))
# .Fortran(ptr, array(1))
```

Lets compile a convolve function.

```{r}

convolve_r <- function(a, b) {
  ab = double(length(a) + length(b) - 1)
  for (i in seq_along(a))
    for (j in seq_along(b))
      ab[i + j - 1] = ab[i + j - 1] + a[i] * b[j]
    ab
}

convolve_rf <- function(a, b, ab) {
  manifest(
    Var(i, "i"),
    Var(j, "i"),
    Var(a, "d", 1, modifiable = FALSE),
    Var(b, "d", 1, modifiable = FALSE),
    Var(ab, "d", 1, modifiable = TRUE)
  )
  for (i in seq_along(a))
    for (j in seq_along(b))
      ab[i + j - 1] = ab[i + j - 1] + a[i] * b[j]
}
```

Here is what it looks like translated to Fortran:
```{r}
frast:::translate(convolve_rf)
```


and also compare it to some alternative approaches to see what the speedup is.




```{r}
convolve_c_ <- inline::cfunction(
  signature(
    a = "double",
    na = "integer",
    b = "double",
    nb = "integer",
    ab = "double"
  ),
  body = "
//void convolve(double *a, int *na, double *b, int *nb, double *ab)
{
    int nab = *na + *nb - 1;

    for(int i = 0; i < nab; i++)
        ab[i] = 0.0;
    for(int i = 0; i < *na; i++)
        for(int j = 0; j < *nb; j++)
            ab[i + j] += a[i] * b[j];
}", convention = ".C")

convolve_c <- function(a, b) {
  convolve_c_(a, length(a), b, length(b), 
              double(length(a) + length(b) - 1))[[5L]]
}

convolve_r <- function(a, b) {
  ab = double(length(a) + length(b) - 1)
  for (i in seq_along(a))
    for (j in seq_along(b))
      ab[i + j - 1] = ab[i + j - 1] + a[i] * b[j]
    ab
}

convolve_rf <- function(a, b, ab) {
  manifest(
    Var(i, "i"),
    Var(j, "i"),
    Var(a, "d", 1, modifiable = FALSE),
    Var(b, "d", 1, modifiable = FALSE),
    Var(ab, "d", 1, modifiable = TRUE)
  )
  for (i in seq_along(a))
    for (j in seq_along(b))
      ab[i + j - 1] = ab[i + j - 1] + a[i] * b[j]
}

ptr <- load_so(convolve_rf)

convolve_f <- function(a, b) {
  .ModernFortran(ptr, a, b, double(length(a) + length(b) - 1),
                 DUP = FALSE)[[3L]]
}

a <- runif(1024)
b <- runif(32)
all.equal(convolve_r(a, b), convolve_f(a, b))
all.equal(convolve_r(a, b), convolve_c(a, b))

r <- bench::mark(convolve_f(a, b), 
                 convolve_c(a, b), 
                 convolve_r(a, b))
```


```{r, dev = 'ragg_png'}
plot(r)
```

```{r}
sessioninfo::session_info()
sessioninfo::platform_info()
```

